Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-09-02 paper/ms.tex
Future apps added

\item Possible future applications:\\
  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
      with Lorentzians, not Gaussians.\\
  b). Multi-waveband GPs.\\
  c). RV interpolation.\\
  d). Non-parametric phase functions.

@@

>>>>

\item Possible future applications:\\
  a). Inference of asteroseismic variability - ala Andrew Gordon Wilson \citep{2013arXiv1302.4245W}, but
      with Lorentzians, not Gaussians.\\
  b). Multi-waveband GPs.\\
  c). RV interpolation.\\
  d). Non-parametric phase functions.\\
  e). Better calibration of the flicker log(g) relation.\\
  f). Measurement of rotational periods - gyrochronology.\\
  g). Doppler shifts in oscillating EBs.

@@

Here we adapt the generalized Rybicki-Press (GRP) algorithm to handle periodic kernels.  We
follow closely the notation and equations given in \citet{Ambikasaran2015}.  We start
with the observation that a Lorentzian correlation function may be decomposed into the sum of two
exponential functions:
\begin{eqnarray}
K(\tau) &=& \alpha e^{-\beta_R \tau} \cos{(\omega \tau)}\\
 &=& \frac{1}{2}\alpha \left[e^{-\beta \tau}+e^{-\beta^* \tau}\right],
\end{eqnarray}
where $\beta = \beta_R + i\beta_I$ is the 
complex kernel parameter (i.e.\ $\beta_R = Re(\beta)$, $\beta_I = Im(\beta)$), $\omega = -\beta_I = 2\pi/P$
is the frequency of oscillation of the kernel with period $P$, and $x^*$ is the complex conjugate of $x$.

Now, the Lorentzian kernal may be included in the GRP algorithm 
using the two exponential kernels with $\beta_1 = \beta$ and $\beta_2 = \beta^*$.  This 
requires the algorithm to use complex arithmetic, which has several disadvantages:
1).\ complex arithmetic is more expensive by a factor of 2-3; 2).\ the complex conjugate equation
is redundant as it turns out that $l_2 = l_1^*$, $r_2 = r_1^*$, and $x_i = x_i^*$ (i.e.\ $x_i$ is
real), meaning that the same equations are being solved twice for the real and complex components; 
and 3).\ if the derivative of the likelihood is computed with automatic
differentiation (AD), most AD algorithms require real variables.

Given these drawbacks, we instead modify the GRP algorithm to use a combination of two real
components rather than two complex components;  this addresses all three of these drawbacks
simultaneously.  The one cost is that the banded extended matrix has a bandwidth which is
larger by one, both above and below the diagonal.

Rather than using $\beta$ and its complex conjugate as components of the kernel, we instead
find the real and imaginary components of the complex equationi for $\beta$, and 
discard the $\beta^*$ equation, which is redundant.  This gives four additional equations for
each $\beta_j$ with non-zero imaginary component, while the kernel components with real
$\beta_j$ still only have two additional components in the extended matrix.  We will enumerate
the number of real $\beta$'s as $p_0$, while the total number of $\beta$'s is still $p$.
This gives a total number of equations to be solved with the extended matrix of $N_{ex}= 
(4(p-p_0)+2p_0+1)\times(N-1)+1$, where $n$ is the number of data points.

For the complex values of $\beta$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric 
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $p$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,p\}$.  Note that the first $p_0$ elements we take
to be the real values of $\beta$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-\beta_1 t_{k,k+1})} ... \exp{(-\beta_p t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-\beta_j t_{k,k+1}}
\cos{(\beta_j^I t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-\beta_j t_{k,k+1}}\sin{(\beta_j^I t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation}
% See 9/2/2016 notes
\sum_{j=1}^p \frac{\alpha_i}{2} l_{k,j}^R + \frac{d}{2} x_k + \sum_{j=1}^p \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{b_k}{2},
\end{equation}
for $j \in \{1,...,p\}$ and $k \in \{1,...,N\}$.  There is no imaginary component to equation (61).

Equation (60) has real and imaginary components:
\begin{eqnarray}
\gamma_{k,j}^R l_{k,j}^R +\gamma_{k,j}^I l_{k,j}^I + \gamma_{k,j}^R x_k - l_{k+1,j}^R &=& 0,\\
\gamma_{k,j}^I l_{k,j}^R -\gamma_{k,j}^R l_{k,j}^I + \gamma_{k,j}^I x_k + l_{k+1,j}^I &=& 0,
\end{eqnarray}
with $j \in \{1,...,p\}$ and $k \in \{1,...,N-1\}$
\citep[note that we have shifted the indices of this equation by one from the indexing using in][]{Ambikasaran2015}.

Finally, equation (59) has real and imaginary components:
\begin{align}
-&r_{k,j}^R + \frac{\alpha_j}{2} x_k &+ \gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I &&=0,\\
 &r_{k,j}^I &+ \gamma_{k,j}^I r_{k+1,j}^R - \gamma_{k,j}^R r_{k+1,j}^I &&=0,
\end{align}
with $j \in \{1,...,p\}$ and $k \in \{2,...,N\}$.  Note that for the $p_0$ real components
we can drop the variables $r_{k,j}^I$ and $l_{k,j}^I$ as these are zero, and we can drop the
imaginary components of equations (59) and (60) as well for these, which reduces the total number
of equations to solve.

With these additional equations, we can solve for the likelihood of sums of full Lorentzian kernels 
in $O(N)$ operations, but with slightly more computational expense due to having 4 extra variables 
for each Lorentzian with non-zero $\omega$, and due to having a slightly larger bandwidth due to the 
cross terms between the real and imaginary variables.  The off-diagonal components are $2(p+1)$ above 
and below the diagonal, for a total bandwidth of $4p+5$; in the case in which $p_0=p$ so that the
imaginary components of $\beta$ are zero, the off-diagonal components revert to $p+1$, giving
a bandwidth of $2p+3$.

As with the GRP method, we can buid an extended matrix which is real and symmetric, and which may be 
solved with a banded solver, as in \citet{NumericalRecipes}.  The determinant of this extended matrix 
may be computed from the LU decomposition, and it is related to the determinant of the kernel by a 
factor of $2^N$ (this is due to dividing each equation by $2$ to yield a symmetric matrix).
We arrange the auxiliary variables with real values of 
$\beta$ first, followed by the complex variables, yielding the vector
$x_{ex} = [x_1, \{r_{2,j}^R,l_{2,j}^R;j=1,...,p_0\}, \{r_{2,j}^R,r_{2,j}^I,l_{2,j}^R,l_{2,j}^I;j=p_0+1,...,p\},
x_2, ..., x_N]$.

{\bf Should I give some examples of this?  Some pseudocode? Some colored matrices like Figures 1-2?}
